---
title: "Positive Only Analysis"
author: "Priscila Analysis 2025"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "/home/chanyue/Single_Cell/Priscila_Analysis_2025/")
```

## Setup

```{r libraries}
library(Seurat)
library(dplyr)
library(ggplot2)
library(harmony)
```

## Load data

```{r load-data}
load("./single_cell_object_harmony_w_sampleNames.RData")
sample
sample$Sample
```

## Treg marker dot plot (positive vs negative)

```{r treg-dotplot}
genes <- c(
  "FOXP3","LAG3","CTLA4","LRRC32","IKZF2",
  "TIGIT","IL2RA","RTKN2","ENTPD1","IL1R2","IL4I1"
)

p = DotPlot(
  sample,
  features = genes,
  group.by = "Sample"  # column in sample@meta.data
) + RotatedAxis()
p
ggsave("./Treg_marker_pos_vs_neg_dotplot.png", plot = p, width = 8, height = 4)
```

## Th17 marker dot plot (positive vs negative)

```{r th17-dotplot}
genes <- c(
  "RORA", "RORC", "KLRB1",
  "IL7R", "IL17A", "CCL20", "IL23A", "IL22", "IL1R1", "CSF2", "IL26",
  "RBPJ", "IL17F", "CCR4", "CCR6"
)

p = DotPlot(
  sample,
  features = genes,
  group.by = "Sample"  # column in sample@meta.data
) + RotatedAxis()
p
ggsave("./Th17_marker_pos_vs_neg_dotplot.png", plot = p, width = 8, height = 4)
```

## Subset positives and re-process

```{r subset-positives}
Idents(sample) = sample$Sample
pos_idents <- grep("_neg", Idents(sample), value = TRUE, invert = TRUE)
positives <- subset(sample, idents = pos_idents)
positives@project.name = "Th17_Treg_Project_Positive"
UpdateSeuratObject(positives)

positives = NormalizeData(positives, verbose = FALSE)
positives = FindVariableFeatures(positives, selection.method = "vst", nfeatures = 3000)
positives = ScaleData(positives, verbose = FALSE)
positives = RunPCA(positives, verbose = TRUE)
positives = RunHarmony(positives, assay.use="RNA", group.by.vars = "Sample", dims.use = 1:30)
positives <- RunUMAP(positives, reduction = "harmony", dims = 1:30)
positives = FindNeighbors(positives, reduction = "harmony", dims = 1:30)
positives = FindClusters(positives, resolution = c(0.2, 0.4, 0.6, 0.8, 1))
```

## Cluster tree

```{r clustree}
library(clustree)
p_tree <- clustree(positives, prefix = "RNA_snn_res.",
                   prop_filter = 0.2)
p_tree
ggsave("./Cluster_Tree.png", plot = p_tree, width = 10, height = 8)
```

## Sample UMAP

```{r sample-umap}
sample_plot = DimPlot(positives, group.by = "Sample", label = TRUE)
sample_plot
ggsave("./sample_UMAP.png", plot = sample_plot, width = 8, height = 6)

save(positives, file = "./positive_combined_harmony_seurat_allPCs.RData")
```

## Feature plots of key markers

```{r feature-plots}
Th17_genes = c("IL2RA","FOXP3","CTLA4","RORA","RBPJ","IL17F","IL22","IL26")
p = FeaturePlot(positives, features = Th17_genes, label = TRUE, pt.size = 0.05, label.size = 3)
ggsave("./FeaturePlot_KeyMarkers.png", plot = p, width = 8, height = 6)
```

## UMAP at resolution 0.8

```{r umap-0.8}
umap = DimPlot(positives, group.by = "RNA_snn_res.0.8", label = TRUE)
umap
ggsave("./UMAP_0.8.png", plot = umap, width = 8, height = 6)
```

## Cluster markers (resolution 0.8)

```{r cluster-markers}
positives.markers <- FindAllMarkers(positives, group.by = "RNA_snn_res.0.8", test.use = "wilcox", only.pos = TRUE, min.pct = 0.1, logfc.threshold = 0.25)
positives.markers
positives.markers %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), p_val_adj, .by_group = TRUE) -> positives.markers
write.csv(positives.markers, file = "./cluster_markers_0.8.csv")
```

## Top markers dot plot

```{r top-markers-dotplot}
positives.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 5) %>%
  ungroup() -> top5

top5_genes = unique(as.character(top5$gene))
p = DotPlot(positives, features = top10_genes, group.by = "RNA_snn_res.0.8") + coord_flip()
p
ggsave("./Dotplot_top5.png", plot = p, width = 10, height = 24)
```

## Key marker dot plot

```{r key-markers-dotplot}
genes = c("RORA", "RORC", "RBPJ", "CCL20", "IL22", "IL26", "IL17F",
          "IL17A", "IL23A", "IL21", "TNFRSF18", "IL4I1", "IL1R1", "CTLA4",
          "LAG3", "TIGIT", "IL2RA", "FOXP3")

p = DotPlot(positives, features = genes, group.by = "RNA_snn_res.0.8") + coord_flip()
ggsave("./Dotplot_KeyMarkers.png", plot = p, width = 10, height = 12, bg = "white")
```

## Export markers to Excel

```{r export-excel}
library(readr)
#install.packages("openxlsx")
positives.markers <- positives.markers %>%
  mutate(.cluster_num = parse_number(as.character(cluster))) %>%
  arrange(.cluster_num, desc(avg_log2FC), p_val_adj) %>%
  group_by(cluster)

library(openxlsx)
# 2) Build the sheet-2 gene columns in the correct cluster order
gene_list <- positives.markers %>%
  summarise(genes = list(gene),
            .cluster_num = first(.cluster_num),
            .groups = "drop") %>%
  arrange(.cluster_num)

# Convert to data frame where each cluster's genes are in a separate column
gene_df <- do.call(cbind, lapply(gene_list$genes, function(x) {
  length_diff <- max(lengths(gene_list$genes)) - length(x)
  c(x, rep(NA, length_diff)) # pad with NA for equal length
}))
colnames(gene_df) <- paste0("Cluster_", gene_list$cluster)
gene_df <- as.data.frame(gene_df)
gene_df

positives.markers
# Write to Excel with two sheets
wb <- createWorkbook()
addWorksheet(wb, "All_Markers")
addWorksheet(wb, "Genes_by_Cluster")

writeData(wb, "All_Markers", positives.markers)
writeData(wb, "Genes_by_Cluster", gene_df)

saveWorkbook(wb, file = "./cluster_markers_0.8.xlsx", overwrite = TRUE)
```

## Stacked barplot of cell counts by sample

```{r barplot-sample}
# plot stacked barplot for annotating the clusters
Idents(positives) = positives$RNA_snn_res.0.8
positives$orig.ident = positives$Sample
orig = positives$orig.ident
orig
u_orig = sort(unique(orig))
u_orig
cluster = as.character(Idents(positives))
uct = sort(unique(cluster))
z = matrix(nrow=length(uct), ncol=length(u_orig))
rownames(z) = uct
colnames(z) = u_orig
z
for (i in 1:nrow(z)){
  for (j in 1:ncol(z)){
    z[i,j] = sum(orig[cluster==uct[i]] == u_orig[j])
  }
}
# Make a plot about num cells for each cluster by orig
z = as.matrix(z)
z1 = data.frame(ncells=as.vector(z),
                type=rep(as.integer(rownames(z)), ncol(z)),
                orig=rep(colnames(z), each=nrow(z)))
p = ggplot(z1, aes(x=type, y=ncells, fill=orig, group=orig)) +
  geom_bar(stat="identity", width=0.7) +
  theme_classic() +
  labs(x="Cluster Number", y="Numer of Cells", fill=NULL)+
  scale_x_continuous(breaks=0:max(z1$type))
p
ggsave("./Barplot_Sample.png", plot = p, width = 8, height = 6)
```

## Average expression by cluster

```{r average-expression}
cluster.averages <- AverageExpression(positives, assays = "RNA")
write.csv(cluster.averages, "ave_expr_by_cluster.csv")
```
