---
title: "Treg/Th17 Module Scores Along RNA Velocity Pseudotime"
author: "Chanyue Charlotte Hu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "/home/chanyue/Single_Cell/Priscila_Analysis_2025/Positive_CD3/")
```

## Overview

Takes the CD3-positive Treg/Th17 subset, transfers the scVelo velocity
pseudotime onto it, scores each cell for the Th17 / Transitional / Treg gene
modules, and plots those scores on the UMAP and along pseudotime.

## Setup

```{r libraries}
library(Seurat)
library(scales)   # squish(), for clamping the color scale
library(dplyr)
library(tidyr)
library(ggplot2)
```

All output for this run goes to one dated folder. Bump `out_dir` for a new run
rather than editing paths further down.

```{r output-dir}
out_dir <- "./pseudotime_plot/05_2026"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load the Treg/Th17 object

```{r load-object}
load("./30PCs/new_analysis_03_2026/Treg_Th17s.RData")
```

## Transfer velocity pseudotime onto the Seurat object

The pseudotime CSV comes from the RNA velocity analysis, where cells are named
`<sample>:<barcode>x` (e.g. `D3positive_RNA:AAACCCACACAATGAAx`). Seurat names
them `<barcode>-<sample number>` (e.g. `AAACCCACACAATGAA-2`), so split the
velocity name apart and rebuild it in Seurat's format before matching.

```{r read-pseudotime}
pseudotime <- read.csv(
  "~/Single_Cell/Priscila_D3toD7_RNA_Velocity/figures/2026/new_analysis_03_2026/IL1R1_velocity_pseudotime_info.csv"
)

# Sample is everything before the colon; barcode is between the colon and the
# trailing "x".
pseudotime$sample  <- sub(":.*", "", pseudotime$X)
pseudotime$barcode <- sub(".*:(.*)x$", "\\1", pseudotime$X)

# Sample -> the numeric suffix Seurat gave that sample when the 10 runs were
# merged, in load order (negative/positive alternating from D3 to D7).
suffix_map <- c(
  D3negative_RNA = "-1",
  D3positive_RNA = "-2",
  D4negative_RNA = "-3",
  D4positive_RNA = "-4",
  D5negative_RNA = "-5",
  D5positive_RNA = "-6",
  D6negative_RNA = "-7",
  D6positive_RNA = "-8",
  D7negative_RNA = "-9",
  D7positive_RNA = "-10"
)
pseudotime$cell <- paste0(pseudotime$barcode, suffix_map[pseudotime$sample])

head(pseudotime[, c("X", "sample", "barcode", "cell", "velocity_pseudotime")])
```

```{r add-pseudotime-metadata}
# Named vector lookup: cells with no velocity pseudotime get NA.
vpt <- pseudotime$velocity_pseudotime
names(vpt) <- pseudotime$cell

Treg_Th17$velocity_pseudotime <- vpt[Cells(Treg_Th17)]

# Sanity check on how many cells actually got a pseudotime.
sum(!is.na(Treg_Th17$velocity_pseudotime))
```

## Gene modules

Three modules along the Th17 -> Treg reprogramming axis. Note the modules
deliberately overlap (e.g. `CTLA4` is in both Transitional and Treg).

```{r gene-modules}
# Th17: effector / inflammatory cytokines plus immediate-early TFs.
th17_genes <- c(
  "IL26", "IL17F", "IL22", "IL17A", "EGR1", "FOS", "EGR2",
  "NR4A3", "NR4A1", "NR4A2", "EGR3"
)

# Transitional: reprogramming intermediate.
transitional_genes <- c(
  "IL10", "IRF4", "CTLA4", "LAG3", "IL1R2", "IL1R1"
)

# Treg: suppressive / regulatory program.
treg_genes <- c(
  "CTLA4", "ENTPD1", "TIGIT", "IKZF2", "LAYN", "FOXP3"
)
```

## Score the modules

`AddModuleScore` appends a "1" to each name, so these land in the metadata as
`Th171`, `Treg1`, and `Transition1`.

```{r add-module-scores}
Treg_Th17 <- AddModuleScore(Treg_Th17, features = list(th17_genes),         name = "Th17")
Treg_Th17 <- AddModuleScore(Treg_Th17, features = list(treg_genes),         name = "Treg")
Treg_Th17 <- AddModuleScore(Treg_Th17, features = list(transitional_genes), name = "Transition")
```

## Label cell types from the 0.8-resolution clusters

Used to color the per-gene plots below.

```{r assign-celltype}
Treg_Th17$celltype <- case_when(
  as.character(Treg_Th17$RNA_snn_res.0.8) %in% c("10", "1") ~ "Th17",
  as.character(Treg_Th17$RNA_snn_res.0.8) == "0"            ~ "Transitional",
  as.character(Treg_Th17$RNA_snn_res.0.8) %in% c("9", "6")  ~ "Treg",
  TRUE                                                      ~ "Other"
)

table(Treg_Th17$celltype)
```

Shared cell type colors, so every plot below uses the same key.

```{r celltype-colors}
celltype_colors <- c(
  "Th17"         = "#4E79A7",  # blue
  "Transitional" = "#59A14F",  # green
  "Treg"         = "#F28E2B",  # orange
  "Other"        = "grey70"
)
```

## Cell density along pseudotime

Where each cluster sits along the trajectory, ignoring module scores.

```{r density-plot}
df <- Treg_Th17@meta.data

p <- ggplot(df, aes(x = velocity_pseudotime, color = RNA_snn_res.0.8, fill = RNA_snn_res.0.8)) +
  geom_density(alpha = 0.3, adjust = 1) +
  labs(
    x = "Velocity pseudotime",
    y = "Cell density",
    title = "Cell density along pseudotime"
  ) +
  theme_classic()

p
ggsave(file.path(out_dir, "Cell_count_density_no_celltype.png"),
       plot = p, width = 8, height = 6, dpi = 600)
```

## Module scores on the UMAP

The three UMAP plots are identical apart from which score they color by, so
they share one theme and one plotting helper.

```{r umap-theme}
theme_score_umap <- function() {
  theme(
    panel.background = element_blank(),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "black"),
    axis.text.x      = element_text(size = 18, color = "black"),
    axis.text.y      = element_text(size = 18, color = "black"),
    axis.title.x     = element_text(size = 18, color = "black"),
    axis.title.y     = element_text(size = 18, color = "black"),
    legend.title     = element_text(size = 18, color = "black"),
    legend.text      = element_text(size = 18, color = "black"),
    plot.title       = element_text(size = 20, color = "black", hjust = 0.5)
  )
}
```

```{r umap-score-helper}
umap_df <- as.data.frame(Embeddings(Treg_Th17, "umap"))
umap_df$Th17_score       <- Treg_Th17$Th171
umap_df$Treg_score       <- Treg_Th17$Treg1
umap_df$Transition_score <- Treg_Th17$Transition1

# Scale is clamped to [0, 0.51]: negative scores are floored to grey (they are
# not biologically meaningful here) and squish() keeps out-of-range highs at the
# top color instead of dropping those cells.
plot_score_umap <- function(score_col, legend_title, plot_title) {
  ggplot(umap_df, aes(x = umap_1, y = umap_2)) +
    geom_point(aes(color = .data[[score_col]]), alpha = 0.6, size = 0.8) +
    scale_color_gradient(
      low    = "grey70",
      high   = "red",
      limits = c(0, 0.51),
      oob    = scales::squish
    ) +
    labs(x = "UMAP 1", y = "UMAP 2", color = legend_title, title = plot_title) +
    theme_score_umap()
}
```

```{r umap-th17}
p <- plot_score_umap("Th17_score", "Th17 score", "Th17 module score on UMAP")
p
ggsave(file.path(out_dir, "module_score_Th17_featureplot.png"),
       plot = p, width = 10, height = 6, dpi = 600)
```

```{r umap-treg}
p <- plot_score_umap("Treg_score", "Treg score", "Treg module score on UMAP")
p
ggsave(file.path(out_dir, "module_score_Treg_featureplot.png"),
       plot = p, width = 10, height = 6, dpi = 600)
```

```{r umap-transitional}
p <- plot_score_umap("Transition_score", "Transitional score", "Transitional module score on UMAP")
p
ggsave(file.path(out_dir, "module_score_Transitional_featureplot.png"),
       plot = p, width = 10, height = 6, dpi = 600)
```

## All three module scores along pseudotime

The main trend figure: one loess curve per module over pseudotime.

```{r module-trend-plot}
score_long <- data.frame(
  velocity_pseudotime = Treg_Th17$velocity_pseudotime,
  Th17_score          = Treg_Th17$Th171,
  Treg_score          = Treg_Th17$Treg1,
  Transition_score    = Treg_Th17$Transition1
) %>%
  pivot_longer(
    cols      = c(Th17_score, Treg_score, Transition_score),
    names_to  = "module",
    values_to = "score"
  )

p <- ggplot(score_long, aes(x = velocity_pseudotime, y = score, color = module)) +
  geom_point(alpha = 0.15, size = 0.4) +
  geom_smooth(se = FALSE, method = "loess", span = 0.5, linewidth = 1.5) +
  scale_color_manual(
    values = c(
      "Th17_score"       = "#4E79A7",
      "Treg_score"       = "#F28E2B",
      "Transition_score" = "#59A14F"
    ),
    labels = c(
      "Th17_score"       = "Th17",
      "Treg_score"       = "Treg",
      "Transition_score" = "Transition"
    )
  ) +
  labs(
    x     = "Velocity pseudotime",
    y     = "Module score",
    title = "Module scores along velocity pseudotime",
    color = "Module"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    axis.text.x  = element_text(size = 13, face = "bold"),
    axis.text.y  = element_text(size = 13, face = "bold")
  )

p
ggsave(file.path(out_dir, "module_score_trend_plot_1.png"),
       plot = p, width = 10, height = 6, dpi = 600)
```

## Per-gene expression along pseudotime

One PNG per module gene, colored by cell type with a single black loess trend
across all cells. `unique()` because the modules share genes.

```{r fetch-gene-expression}
genes <- unique(c(th17_genes, transitional_genes, treg_genes))

plot_long <- FetchData(
  Treg_Th17,
  vars  = c("velocity_pseudotime", "celltype", genes),
  layer = "data"
) %>%
  filter(!is.na(velocity_pseudotime)) %>%
  pivot_longer(
    cols      = all_of(genes),
    names_to  = "gene",
    values_to = "expression"
  )
```

```{r per-gene-plots}
gene_dir <- file.path(out_dir, "individual_genes")
dir.create(gene_dir, recursive = TRUE, showWarnings = FALSE)

for (curr_gene in genes) {

  curr_df <- plot_long %>% filter(gene == curr_gene)

  p <- ggplot(curr_df, aes(x = velocity_pseudotime, y = expression, color = celltype)) +
    geom_point(alpha = 0.25, size = 0.4) +
    geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.8, span = 0.75) +
    scale_color_manual(values = celltype_colors) +
    # Points are tiny and transparent; bump them up in the legend only.
    guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
    labs(
      x     = "RNA Velocity Pseudotime",
      y     = "Gene expression",
      color = "Cell Type",
      title = paste0(curr_gene, " Expression Along Velocity Pseudotime")
    ) +
    theme_classic(base_size = 12) +
    theme(
      axis.text    = element_text(size = 10, color = "black"),
      axis.title   = element_text(size = 12, color = "black"),
      legend.title = element_text(size = 12, color = "black"),
      legend.text  = element_text(size = 11, color = "black"),
      plot.title   = element_text(size = 12, hjust = 0.5, face = "bold")
    )

  ggsave(
    filename = file.path(gene_dir, paste0(curr_gene, "_pseudotime_plot.png")),
    plot     = p,
    width    = 6,
    height   = 4,
    dpi      = 600
  )
}
```

## Exploratory
<!-- Not saved to disk — kept for interactive checks. -->

<!-- Th17 score against pseudotime by cluster, zoomed into the second half of the -->
<!-- trajectory where the Treg shift happens. -->

<!-- ```{r explore-th17-zoom, eval=FALSE} -->
<!-- ggplot(df, aes(x = velocity_pseudotime, y = Th171, color = RNA_snn_res.0.8)) + -->
<!--   geom_point(alpha = 0.5, size = 1) + -->
<!--   geom_smooth(method = "loess", color = "grey30", se = FALSE) + -->
<!--   labs( -->
<!--     x     = "RNA Velocity Pseudotime", -->
<!--     y     = "Th17 Module Score", -->
<!--     title = "Th17 Module Score vs. Velocity Pseudotime" -->
<!--   ) + -->
<!--   theme_classic(base_size = 14) + -->
<!--   coord_cartesian(xlim = c(0.5, 1)) -->
<!-- ``` -->

<!-- NLRP3 (not in any of the modules above). -->

<!-- ```{r explore-nlrp3, eval=FALSE} -->
<!-- gene_df <- FetchData(Treg_Th17, vars = c("velocity_pseudotime", "NLRP3")) -->

<!-- ggplot(gene_df, aes(x = velocity_pseudotime, y = NLRP3)) + -->
<!--   geom_point(alpha = 0.15, size = 0.4, color = "#713948") + -->
<!--   geom_smooth(se = TRUE, method = "loess", span = 0.3, linewidth = 1.2, color = "#713948") + -->
<!--   labs( -->
<!--     x     = "Velocity pseudotime", -->
<!--     y     = "NLRP3 expression", -->
<!--     title = "NLRP3 expression along velocity pseudotime" -->
<!--   ) + -->
<!--   theme_classic(base_size = 14) + -->
<!--   theme( -->
<!--     axis.title.x = element_text(size = 16, face = "bold"), -->
<!--     axis.title.y = element_text(size = 16, face = "bold"), -->
<!--     axis.text.x  = element_text(size = 13), -->
<!--     axis.text.y  = element_text(size = 13) -->
<!--   ) -->

<!-- FeaturePlot( -->
<!--   Treg_Th17, -->
<!--   features   = "NLRP3", -->
<!--   reduction  = "umap", -->
<!--   cols       = c("grey70", "red"), -->
<!--   min.cutoff = 0, -->
<!--   max.cutoff = 0.51, -->
<!--   pt.size    = 0.8 -->
<!-- ) + -->
<!--   labs(title = "NLRP3 expression on UMAP", color = "NLRP3 expression") -->
<!-- ``` -->
