---
title: "Positive CD3 Only Analysis"
author: "Chanyue Charlotte Hu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "/home/chanyue/Single_Cell/Priscila_Analysis_2025/Positive_CD3/")
```

## Setup

```{r libraries}
library(Seurat)
library(dplyr)
library(ggplot2)
library(harmony)
```


## Finalize CD3 object (new in April 2026)

Load the all-PC CD3 object, plot clusters, set `seurat_clusters` to the 0.8
resolution, and save the final object. `eval=FALSE` because it writes the
final RData.

```{r finalize-cd3, eval=FALSE}
load("./positive_CD3_seurat_allPCs.RData")

p = DimPlot(pos_cd3, group.by = "seurat_clusters", label = TRUE)
p
ggsave("./30PCs/umap_clusters.png", plot = p, width = 8, height = 6, dpi = 600)
pos_cd3$seurat_clusters = pos_cd3$RNA_snn_res.0.8
save(pos_cd3, file = "./30PCs/positive_CD3_seurat_final.RData")
```

## Subset CD3-positive cells and re-process

```{r subset-cd3, eval=FALSE}
load("../single_cell_object_harmony_w_sampleNames.RData")
unique(sample$Sample)

Idents(sample) = sample$Sample
pos_idents <- grep("CD3_pos", Idents(sample), value = TRUE)
pos_cd3 <- subset(sample, idents = pos_idents)
pos_cd3@project.name = "Th17_Treg_Project_Positive_CD3"
UpdateSeuratObject(pos_cd3)

pos_cd3 = NormalizeData(pos_cd3, verbose = FALSE)
pos_cd3 = FindVariableFeatures(pos_cd3, selection.method = "vst", nfeatures = 3000)
pos_cd3 = ScaleData(pos_cd3, verbose = FALSE)
pos_cd3 = RunPCA(pos_cd3, verbose = TRUE)
ElbowPlot(pos_cd3)
pos_cd3 = FindNeighbors(pos_cd3, reduction = "pca", dims = 1:30)
pos_cd3 = FindClusters(pos_cd3, reduction = "pca", resolution = c(0.2, 0.4, 0.6, 0.8))
pos_cd3 = RunUMAP(pos_cd3, dims = 1:30)
```

## Cluster tree

```{r clustree, eval=FALSE}
library(clustree)
p_tree <- clustree(pos_cd3, prefix = "RNA_snn_res.",
                   prop_filter = 0.2)
p_tree
```

## UMAPs across resolutions (all-PC object)

```{r umaps-allpc}
load("./positive_CD3_seurat_allPCs.RData")
pos_cd3

umap = DimPlot(pos_cd3, group.by = "RNA_snn_res.0.6", label = TRUE)
umap

umap = DimPlot(pos_cd3, group.by = "RNA_snn_res.0.8", label = TRUE)
umap
ggsave("./UMAP_0.8.png", plot = umap, width = 8, height = 6)

umap = DimPlot(pos_cd3, group.by = "RNA_snn_res.0.4", label = TRUE)
umap
```

## Cluster markers (0.8, 30PC — decided in March 2026)) 

```{r markers-allpc-0.8, eval=FALSE}
pos_cd3.markers <- FindAllMarkers(pos_cd3, group.by = "RNA_snn_res.0.8", test.use = "wilcox", only.pos = TRUE, min.pct = 0.1, logfc.threshold = 0.25)
pos_cd3.markers
pos_cd3.markers %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), p_val_adj, .by_group = TRUE) -> pos_cd3.markers
write.csv(pos_cd3.markers, file = "./cluster_markers_0.8.csv")
```

## Top markers dot plot

```{r top-markers-dotplot}
pos_cd3.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 5) %>%
  ungroup() -> top5

top5_genes = unique(as.character(top5$gene))
p = DotPlot(pos_cd3, features = top5_genes, group.by = "RNA_snn_res.0.8") + coord_flip()
p
ggsave("./Dotplot_top5_0.8.png", plot = p, width = 10, height = 24)
```

## Key marker dot plot

```{r key-markers-dotplot}
genes = c("RORA", "RORC", "RBPJ", "CCL20", "IL22", "IL26", "IL17F",
          "IL17A", "IL23A", "IL21", "TNFRSF18", "IL4I1", "IL1R1", "CTLA4",
          "LAG3", "TIGIT", "IL2RA", "FOXP3")

p = DotPlot(pos_cd3, features = genes, group.by = "RNA_snn_res.0.8") + coord_flip()
ggsave("./Dotplot_KeyMarkers_0.8.png", plot = p, width = 10, height = 12, bg = "white")
```

## Export markers to Excel

```{r export-excel}
library(readr)
pos_cd3.markers <- pos_cd3.markers %>%
  mutate(.cluster_num = parse_number(as.character(cluster))) %>%
  arrange(.cluster_num, desc(avg_log2FC), p_val_adj) %>%
  group_by(cluster)

library(openxlsx)
# 2) Build the sheet-2 gene columns in the correct cluster order
gene_list <- pos_cd3.markers %>%
  summarise(genes = list(gene),
            .cluster_num = first(.cluster_num),
            .groups = "drop") %>%
  arrange(.cluster_num)

# Convert to data frame where each cluster's genes are in a separate column
gene_df <- do.call(cbind, lapply(gene_list$genes, function(x) {
  length_diff <- max(lengths(gene_list$genes)) - length(x)
  c(x, rep(NA, length_diff)) # pad with NA for equal length
}))
colnames(gene_df) <- paste0("Cluster_", gene_list$cluster)
gene_df <- as.data.frame(gene_df)
gene_df

pos_cd3.markers
# Write to Excel with two sheets
wb <- createWorkbook()
addWorksheet(wb, "All_Markers")
addWorksheet(wb, "Genes_by_Cluster")

writeData(wb, "All_Markers", pos_cd3.markers)
writeData(wb, "Genes_by_Cluster", gene_df)

saveWorkbook(wb, file = "./cluster_markers_0.8.xlsx", overwrite = TRUE)
```

## Stacked barplot of cell counts by donor

```{r barplot-donor}
# plot stacked barplot for annotating the clusters
Idents(pos_cd3) = pos_cd3$RNA_snn_res.0.8
pos_cd3$orig.ident = pos_cd3$Donor
orig = pos_cd3$orig.ident
orig
u_orig = sort(unique(orig))
u_orig
cluster = as.character(Idents(pos_cd3))
uct = sort(unique(cluster))
z = matrix(nrow=length(uct), ncol=length(u_orig))
rownames(z) = uct
colnames(z) = u_orig
z
for (i in 1:nrow(z)){
  for (j in 1:ncol(z)){
    z[i,j] = sum(orig[cluster==uct[i]] == u_orig[j])
  }
}
# Make a plot about num cells for each cluster by orig
z = as.matrix(z)
z1 = data.frame(ncells=as.vector(z),
                type=rep(as.integer(rownames(z)), ncol(z)),
                orig=rep(colnames(z), each=nrow(z)))
p = ggplot(z1, aes(x=type, y=ncells, fill=orig, group=orig)) +
  geom_bar(stat="identity", width=0.7) +
  theme_classic() +
  labs(x="Cluster Number", y="Numer of Cells", fill=NULL)+
  scale_x_continuous(breaks=0:max(z1$type))
p
ggsave("./Barplot_Donor_0.8.png", plot = p, width = 8, height = 6)

cluster.averages <- AverageExpression(pos_cd3, assays = "RNA")
write.csv(cluster.averages, "ave_expr_by_cluster_0.8.csv")

DimPlot(pos_cd3, group.by = "RNA_snn_res.0.8", label = TRUE) + DimPlot(pos_cd3, group.by = "RNA_snn_res.0.4", label = TRUE)
```

## average expression and separate out Treg_Th17 object

```{r treg-th17-subset, eval=FALSE}
avg_expr <- AverageExpression(
  pos_cd3,
  assays = "RNA",
  group.by = "RNA_snn_res.0.8",
)

avg_expr
write.csv(avg_expr, "./30PCs/new_analysis_03_2026/ave_expr_by_cluster_0.8.csv")

Idents(pos_cd3) = pos_cd3$RNA_snn_res.0.8
Treg_Th17 = subset(pos_cd3, idents = c(0, 1, 6, 9, 10))
unique(Treg_Th17$RNA_snn_res.0.8)

plot = DimPlot(Treg_Th17, label = TRUE)
plot
ggsave(plot = plot, filename = "./30PCs/new_analysis_03_2026/cell_selector.png", width = 8, height = 6)
select.cells <- CellSelector(plot = plot)

Treg_Th17 = subset(Treg_Th17, cells = select.cells)
plot = DimPlot(Treg_Th17)
select.cells <- CellSelector(plot = plot)
Treg_Th17 = subset(Treg_Th17, cells = select.cells, invert = TRUE)

save(Treg_Th17, file = "./30PCs/Treg_Th17s.RData")
```
